### 2020-05-28 paper.tex
Finish Wrap Up

Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs. E.g. when an engineer wants to know which part of the program called a particular function, the stack trace gives a clear answer to this question. Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a stack trace: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It has the ability to push (``emit'') values to its dependents. Multiple depending Observables form and describe a data-flow graph eventually. A multitude of operators allow the transformation of values and (higher-order) Observables. Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

@@

After we built a first intuition for how software engineers work with RP by evaluating the informal interviews, we were interested in more RxJS-specific, hands-on experiences. We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request. After reaching out, we were able to collect five responses: One by a RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS which includes the author of this paper.

The reports provided us diverse perspectives on various areas of software development using RxJS. One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it. I really disliked that once you started to test async code you were forced to pass a TestScheduler''. Allowing to pass the scheduler explicitly as parameter forced them to introduce code which was only necessary for testing reasons they stated further. With its current major version, RxJS 6 improved heavily on the  \mintinline{TypeScript}{TestScheduler}. The runtime environment itself can now be augmented with the scheduler which results in cleaner code.

Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners. One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators. On top, each data source could re-emit updated versions of previously requested information any time. Though tested thoroughly, after a week in production the first of many bugs was reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all'' they told us. The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints. ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy'', hence manually augmenting the code with trace logs. After the time consuming log analysis, they finally were able resolve the bug. The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future. (We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion} more in-depth.)

@@

Most participants considered a statically typed language like TypeScript a fundamental necessity which allowed them to implement data-flow graphs with a minimal, formal correctness. We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and lacking ``the right tool'' to work within the context of RP.

% TODO

The interpretation of the combined interview and war stories data gives us an idea about two main challenges software engineers have to overcome during debugging activities of RxJS-based code:

It is crucial to the engineers to \textbf{understand data flows}  within the Observable graph. Questions like ``Which source emitted a value at what time?'' or ``What is the result of a specific transformation?'' are common in this context. Once a program fails, software engineers need to \textbf{rely on traditional debugger tools}. Though providing help to some extent, the tools lack of specific knowledge about the RP runtime environment adds some oddities to the debugging process: The (i) functionalities allowing step-wise navigation through the code does not respect the RP context. Assume we have set a breakpoint at \mintinline{TypeScript}{i * 2} on line~2 of Listing~\ref{lst:rxjs-step-controls}. Once the debugger hits the breakpoint and we instruct it to ``Step Over'', the debugger will lead us to the internal implementation of the \mintinline{TypeScript}{map} operator instead to line~3. (ii) Stack traces include information about the invocation of internal functions which, most of the time, do not provide any valuable information to the engineers further.

\begin{listing}[H]
	\begin{minted}{TypeScript}
		interval(1000).pipe(      // Increasing numbers with 1s delay
		  map(i => i * 2),        // Breakpoint on i * 2
		  map(i => i + 1),
		  take(4)
		).subscribe(console.log); // 1, 3, 5, 7 & complete
	\end{minted}
	\caption{RxJS Observable With Breakpoint Within a \mintinline{TypeScript}{map} Operator}
	\label{lst:rxjs-step-controls}
\end{listing}

>>>>

Debuggers available in IDEs and the developer tools of a web browser\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} today are specialized to work with imperative, control-flow oriented programs. E.g. when an engineer wants to know which part of the program called a particular function, the call stack gives a clear answer to this question. Assuming a data-flow oriented program implemented using RP, the question, which transformation was executed right before another cannot be answered using a call stack: A breakpoint within the transformation will not point to its logical predecessor, it will lead to the internals of the RP runtime instead.

@@

``Reactive Extensions'' (\emph{ReactiveX}) is an open source project. Its members and contributors defined a generic description of an RP API. They further provide reference implementations of this API along with RP runtime environments for various programming languages like Java, C\# and JavaScript\footnote{\url{http://reactivex.io/languages.html}}. ReactiveX summarizes the API as ``\dots a combination of the best ideas from the Observer pattern, the Iterator pattern, and functional programming''\cite{reactivex}. The core concept of the API specification is the Observable\footnote{There is no known relation between ReactiveX' concept of the Observable and the deprecated Java class \href{https://docs.oracle.com/en/java/javase/11/docs/api/java.base/java/util/Observable.html}{\mintinline{TypeScript}{java.util.Observable}}.}: An Observable can be composed and linked (``piped'') with other Observables. It has the ability to push (``emit'') values to its dependents. Multiple depending Observables form and describe a data-flow graph eventually. A multitude of operators allow the transformation of values and (higher-order\footnote{An Observable emitted by another Observable is considered a Higher-Order Observable. This naming is related to the concept of higher-order functions in mathematics and computer science.}) Observables. Values emitted from an Observable can be subscribed to at any point which is closely related to the Observer design patterns \mintinline{Java}{attach} method.

@@

After we built a first intuition for how software engineers work with RP by evaluating the informal interviews, we were interested in more RxJS-specific, hands-on experiences. We asked the engineering community via Twitter\footnote{\url{https://twitter.com/swissmanu/status/1242429409208029185}} about their personal, most recent RxJS debugging war story and sent out various emails with the same request. After reaching out, we were able to collect five responses: One by an RxJS core team member, two from Angular Google Developer Experts and another two reports by two software engineers building web and mobile applications using React and RxJS which includes the author of this paper.

The reports provided us with diverse perspectives on various areas of software development using RxJS. One of the reporters focused on how they built code which had improved on testability over the recent versions of RxJS: ``[\dots] in the beginning, it was very hard to write async tests and I did not like it. I really disliked that once you started to test async code you were forced to pass a TestScheduler''. Allowing to pass the scheduler explicitly as parameter forced them to introduce code which was only necessary for testing reasons they stated further. With its current major version, RxJS 6 improved heavily on the  \mintinline{TypeScript}{TestScheduler}. The runtime environment itself can now be augmented with the scheduler which results in cleaner code.

Even though the share of non-productive, testing-related code necessary to build mature RxJS-based applications was mentioned to have decreased today, multiple of our reporters mentioned the common practice of manually modifying production code during debugging sessions, hence confirming earlier statements from our informal interview partners. One engineer described a specific scenario where they suspected a problem within a complex Observable composition: Having multiple asynchronous, remote data sources, they used Observables to model the dependencies between them and implemented computations on their results as operators. On top, each data source could re-emit updated versions of previously requested information any time. Though tested thoroughly, after a week in production the first of many bugs was reported: ``Figures were changing multiple times (where I expected them to change once), other numbers were not updated at all'' they told us. The browsers debugger tools and its breakpoints did not help much since the operators were executed several multiple times and some parts of the stream were impossible to grasp, even with conditional breakpoints. ``I started to inspect the flow [\dots] with console.logs and later also using tags from rxjs-spy which exposed more detailed life cycle information'', hence manually augmenting the code with trace logs. After the time consuming log analysis, they finally were able resolve the bug. The log statements added were removed in the aftermath, though the rxjs-spy related code was left in the Observable stream in case they might be needed again in the future. (We will elaborate on the functionality of rxjs-spy in Section~\ref{sec:discussion} more in-depth.)

@@

Most participants considered a statically typed language like TypeScript a fundamental necessity allowing them to implement data-flow graphs with minimal, formal correctness. We noticed throughout all reports that once the engineers were required to interact with the Observable graphs at runtime, they felt kind of ``lost'' and ``lacking the right tool'' to work within the context of RP though. Yet traditional debugging tools allowed them to inspect various parts of a stream, they were unable to get more information about e.g. life cycles of the Observable composition. Almost all of them referred to the practice of modifying their code manually and adding log statements where they assumed a problem. Listing~\ref{lst:rxjs-debugging} exemplifies challenges when debugging a stream of Observables with traditional debugging tools.

\begin{listing}[H]
	\begin{minted}{TypeScript}
		interval(1000).pipe(    // Increasing numbers with 1s delay
		  map(i => i * 2),      // Breakpoint possible within i * 2
		  take(4),              // No breakpoint possible
		  tap(console.log)      // Manually added log statement
		).subscribe(showValue); // 0, 2, 4, 6
	\end{minted}
	\caption{Debugging of an RxJS Observable using breakpoints and log statements. \mintinline{TypeScript}{showValue} renders an emitted value to the UI.}
	\label{lst:rxjs-debugging}
\end{listing}

Where a breakpoint can easily be added to Line~2 within the arrow function, this is impossible for \mintinline{TypeScript}{take} on Line~3. One would need to place the breakpoint within the operators internal implementation instead which can be cumbersome in case the operator is used in a different stream as well. Once the breakpoint on Line~2 interrupts the programs execution, we will notice another shortcoming related to this circumstance: Rather than representing the logical flow implemented using the DSL, the call stack as shown in Listing~\ref{lst:rxjs-call-stack} points deep into RxJS' internal implementation.

\begin{listing}[H]
	\begin{minted}{TypeScript}
		<anonymous> RxJS
			rxjs 6.5.2/internal/operators/map.js:49
			rxjs 6.5.2/internal/Subscriber.js:66
			rxjs 6.5.2/internal/observable/interval.js:23
			rxjs 6.5.2/internal/scheduler/AsyncAction.js:71
			// ...
	\end{minted}
	\caption{Call Stack for Arrow Function on Line~2 in Listing~\ref{lst:rxjs-debugging}}
	\label{lst:rxjs-call-stack}
\end{listing}

We further learned that log statements as shown on Line~4 in Listing~\ref{lst:rxjs-debugging} indeed help to trace emitted values, but do not expose any information about the life cycle of an Observable. The reporter describing their usage of rxjs-spy explained, that the knowledge about when an Observable gets subscribed, unsubscribed or completed helped a lot in order to solve problems related to higher-order Observables.

Finally, we understood that if a problem is hard to replicate within the actual application, the engineers use external sandbox development environments to isolate specific parts of an Observable composition. This allows them to iterate on it faster than it would be possible otherwise.

After the evaluation of all reports and interviews it is our speculation that software engineers are truly lacking debugging tools which ``understand'' RP concepts provided by RxJS. Though traditional debuggers can help up to some extent, they currently cannot answer all questions an engineer has as we saw repeatedly hence making manual modifications on source code necessary. We will validate this assumption after we took a look on previous efforts on the area of reactive debugging.
